\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-27 paper.tex
Yans updates

\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG).
% Be specific about what we're talking about.
ARGs are classically defined by the graph of
% We can play a bit loose here, please don't change to "common ancestor"
coalescence and recombination events in the history of sample,
with the differential inheritance structure encoded by the
crossover breakpoints associated with recombination events.
We show that this ``event ARG'' (eARG) encoding
is limited in the patterns of genetic inheritance that can be represented,
being insufficient to describe the outputs of modern ARG simulation
and inference methods,
and ill-suited to capturing the complex patterns of
relatedness that exist in contemporary datasets.
We propose an alternative ``genome ARG'' (gARG) encoding
expressed in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance.
We show it can fully capture the richness of modern large-scale datasets
and is compatible with modern simulation and inference methods.
Moreover, we show that, unlike the eARG encoding which requires
a precise estimate of all recombination events,
the gARG encoding enables fine-grained
levels of precision about recombination to be represented,
opening up a rich set of additional details.
\end{abstract}

@@

An ARG node with one child and one or
more parents is ARG-unary.
If it has one child and only one parent, it represents a ``pass-through'' node:
these occur where we wish the record the passage of ancestral material
through a known node. For example, in simulations
it is sometimes useful to record the passage of
ancestral material through known pedigree individuals
regardless of whether common ancestry occurs.
Nodes with one child and two parents arise when
we model a recombination event using a single node
in the classical manner
(e.g.\ Fig.~\ref{fig-ancestry-resolution}).
It is also possible for sample nodes to be ARG-unary, for example
in inferences from longitudinal datasets where genetic
data is sampled at many timepoints and recombination
is rare (e.g.~SARS-CoV-2; see Discussion).

Locally unary nodes are a common and important feature
of many different types of ARG.
As discussed in the previous section,
without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them. As we see in the next two sections, both ARG and locally
unary nodes occur in various scenarios, and are
produced by a range of current inference methods.

@@

The timing, positions, and even the number of recombination events is generally
not possible to infer precisely from genome sequencing data. Under
coalescent-based models, the proportion of recombination events that change the
ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
events only a small proportion are actually detectable from the data, assuming
human-like mutation and recombination rates \citep{myers2002detection,hayman2023recoverability}.
Even when a recombination event \emph{is} detectable, its timing and breakpoint
position can only be inferred approximately, depending on how much information
can be elucidated from mutations in the surrounding genomic region.
% For general
% ARGs, for instance those representing pedigrees, or genealogies of organisms with
% low mutation rates and complex patterns of recombination, such inference can be
% even more challenging.
The fact that the eARG encoding \emph{requires}
precise information about recombination is therefore a fundamental limitation.
% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such precise estimates provide
for downstream applications.
Many applications work by examining local trees independently.
For example, the \relate\ selection test~\citep{speidel2019method}
obtains $p$-values by computing clade size probabilities conditional
on the timing of coalescence events in a given local tree.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\cite{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
Similarly, \cite{fan2023likelihood} compute the likelihood
of an ARG under a particular demographic model as the product
over a sample of widely-separated local trees, assumed to be independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account.
Thus, while SIA takes advantage of information about local tree correlation,
it is in quite an indirect way, and
clearly much of the detail about recombination events in an ARG is lost.
The \texttt{tsdate} algorithm and related naive estimator
of ancestral location explicitly use the gARG encoding
to reason about ancestral nodes~\citep{wohns2022unified}.
The main application for fully precise ARGs thus far has been
to compute a likelihood under the
coalescent~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian,
guo2022recombination},
which currently requires the details of all recombination
events to be known.

@@

\section{Cell lineages and ARGs}
\label{sec-cell-lineages-and-args}
In eukaryotic organisms, ARGs are a result of the cellular processes of mitosis
(which leads to common ancestry and coalescence)
and meiosis (which leads to recombination, including both crossovers and gene conversion).

\begin{figure}
\begin{center}
    \includegraphics[width=0.8\textwidth]{illustrations/cell-lines}
\end{center}
\caption{\label{fig-cell-lines}
Cellular inheritance of a single chromosome in a diploid population.
Individuals (blue) contain diploid cells (white circles enclosing a homologous pair of chromosomes).
For clarity, only two rounds of mitotic germ-line cell division are shown per individual, and
meiosis is not illustrated in detail.
Lines show prospective inheritance paths for all chromosomes. Solid lines show all possible
retrospective ancestry paths for four chosen chromosomes (indicated by square black ``sampling events'')
sampled from 3 diploid individuals ($D_1$, $D_2$, $D_3$) in the current generation.
Ancestral recombination events and coalescence events are shown as red and blue squares respectively.
The ARG paths for the lower arm of the sampled chromosomes (green) is highlighted as a
thick solid line and involves a single recombination event and four coalescence events
(highlighted as deep red and blue squares within individuals $D_5$, $D_{10}$, and $D_{13}$).
ARG lineages also show gametic genomes, contained within shaded circles.
As in Fig.~\ref{fig-arg-in-pedigree}A, inherited regions within the sampled chromosome arm are
shaded by the number of descendant samples.
}
\end{figure}

Fig.~\ref{fig-simplification} shows a schematic of the events (coloured red and blue squares) and
the genomes (chromosome icons) that occur in the cellular germline of a simplified, diploid multicellular
hermaphrodite eukaryote with partially overlapping generations. Here, events are not represented by a single
genome: rather, a genome can be associated with (or ``tag'') an event above or below it.
The gARG representation allows any such genome to be arbitrarily distant in time or number of cell
divisions from the event itself (see Section~\ref{sec-precision}).

>>>>

\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG).
% Be specific about what we're talking about.
ARGs are classically defined by the graph of
% We can play a bit loose here, please don't change to "common ancestor"
coalescence and recombination events in the history of sample,
with the differential inheritance structure encoded by the
crossover breakpoints associated with recombination events.
We show that this ``event ARG'' (eARG) encoding
is limited in the patterns of genetic inheritance that can be represented,
being insufficient to describe the outputs of modern ARG simulation
and inference methods,
and ill-suited to capturing the complex patterns of
relatedness that exist in contemporary datasets.
We propose an alternative ``genome ARG'' (gARG) encoding
expressed in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance.
We show this can fully capture the richness of modern large-scale datasets
and is compatible with modern simulation and inference methods.
Moreover, we show that, unlike the eARG encoding which requires
a precise estimate of all recombination events,
the gARG encoding enables fine-grained
levels of precision about recombination to be represented,
opening up a rich set of additional details.
\end{abstract}

@@

% We get discussion of ARG-unary nodes out of the way first: these are
% relatively easy to understand (but it's worth emphasising that
% eARG-like recombination nodes are ARG-unary, as that might not be clear
By definition, ARG-unary nodes have one child but can have one or
more parents.
A node with one child and only one parent represents a ``pass-through'' node:
these occur where we wish the record the passage of ancestral material
through a known node. For example, in simulations
it is sometimes useful to record the passage of
ancestral material through known pedigree individuals
regardless of whether common ancestry occurs.
Nodes with one child and two parents arise when
we model a recombination event using a single node
in the classical manner
(e.g.\ Fig.~\ref{fig-ancestry-resolution}).
It is also possible for sample nodes to be ARG-unary, for example
in inferences from longitudinal datasets where genetic
data is sampled at many timepoints and recombination
is rare (e.g.~SARS-CoV-2; see Discussion).

% Moving on from ARG-unary to the more general case
More generally, locally unary nodes,
which can have one or more children in the graph,
are a common and important feature
of many different types of ARG.
As discussed in the previous section,
without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them. As we see in the next two sections, both ARG and locally
unary nodes occur in various scenarios, and are
produced by a range of current inference methods.

@@

The timing, positions, and even the number of recombination events is generally
not possible to infer precisely from genome sequencing data. Under
coalescent-based models, the proportion of recombination events that change the
ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
events only a small proportion are actually detectable from the data, assuming
human-like mutation and recombination rates \citep{myers2002detection,hayman2023recoverability}.
Even when a recombination event \emph{is} detectable, its timing and breakpoint
position can only be inferred approximately, depending on how much information
can be elucidated from mutations in the surrounding genomic region.
% For general
% ARGs, for instance those representing pedigrees, or genealogies of organisms with
% low mutation rates and complex patterns of recombination, such inference can be
% even more challenging.
The fact that the eARG encoding \emph{requires}
precise information about recombination is therefore a fundamental limitation.
% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such precise estimates provide
for downstream applications.
Many applications work by examining local trees independently.
For example, the \relate\ selection test~\citep{speidel2019method}
obtains $p$-values by computing clade size probabilities conditional
on the timing of coalescence events in a given local tree.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\cite{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
Similarly, \cite{fan2023likelihood} compute the likelihood
of an ARG under a particular demographic model as the product
over a sample of widely-separated local trees, assumed to be independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account.
Thus, while SIA takes advantage of information about local tree correlation,
it is in quite an indirect way, and
clearly much of the detail about recombination events in an ARG is lost.
% You don't even need to define event nodes when using a graph-based
% algorithm like tsdate
The \texttt{tsdate} algorithm and related naive estimator
of ancestral location explicitly use the gARG encoding
to reason about ancestral nodes~\citep{wohns2022unified}.
The main application for fully precise ARGs thus far has been
to compute a likelihood under the
coalescent~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian,
guo2022recombination},
which currently requires the details of all recombination
events to be known.

@@

\section{Cell lineages and ARGs}
\label{sec-cell-lineages-and-args}
\begin{figure}
\begin{center}
    \includegraphics[width=0.8\textwidth]{illustrations/cell-lines}
\end{center}
\caption{\label{fig-cell-lines}
Cellular inheritance of a single chromosome in a diploid population.
Individuals (blue) contain diploid cells (white circles enclosing a homologous pair of chromosomes).
For clarity, only two rounds of mitotic germ-line cell division are shown per individual, and
meiosis is not illustrated in detail.
Lines show prospective inheritance paths for all chromosomes. Solid lines show all possible
retrospective ancestry paths for four chosen chromosomes (indicated by square black ``sampling events'')
sampled from 3 diploid individuals ($D_1$, $D_2$, $D_3$) in the current generation.
Ancestral recombination events and coalescence events are shown as red and blue squares respectively.
The ARG paths for the lower arm of the sampled chromosomes (green) is highlighted as a
thick solid line and involves a single recombination event and four coalescence events
(highlighted as deep red and blue squares within individuals $D_5$, $D_{10}$, and $D_{13}$).
ARG lineages also show gametic genomes, contained within shaded circles.
As in Fig.~\ref{fig-arg-in-pedigree}A, inherited regions within the sampled chromosome arm are
shaded by the number of descendant samples.
}
\end{figure}

In eukaryotic organisms, ARGs are a result of the cellular processes of mitosis
(which leads to common ancestry and coalescence)
and meiosis (which leads to recombination, including both crossovers and gene conversion).
Fig.~\ref{fig-simplification} shows a schematic of the events (coloured red and blue squares) and
the genomes (chromosome icons) that occur in the cellular germline of a simplified, diploid multicellular
hermaphrodite eukaryote with partially overlapping generations. 
% It may be worth putting in something here about events being a slippery concept
% (we have already mentioned this at the end of Section 3
Here, an event is not represented by a specific
genome: rather, a genome can be associated with (or ``tag'') an event above or below it.
The gARG representation allows any such genome to be arbitrarily distant in time or number of cell
divisions from the event itself (see Section~\ref{sec-precision}).
